---
title: "Insert title here"
subtitle: "BCon 147: special topics"
author: "Insert your name here"
date: last-modified
format: 
  html:
    css: custom.scss
    self-contained: true
    mainfont: arial
    fontsize: 1em
    linestretch: 1.5
    toc: true
    toc-location: left
    number-sections: true
    code-fold: false
    theme:
      light: cosmo
      dark: darkly
execute:
  message: false
  warning: false
  fig-align: center
editor: visual
---

```{r}
#| label: setup
#| include: false
#| echo: false

# load your libraries here
library(tidyverse)
library(readxl)
library(janitor)
library(tidytext)

# additional options
## increase max print
getOption("max.print")

```

## Project overiew

In this project, we will explore employee attrition and performance using the HR Analytics Employee Attrition & Performance dataset. The primary goal is to develop insights into the factors that contribute to employee attrition. By analyzing a range of factors, including demographic data, job satisfaction, work-life balance, and job role, we aim to help businesses identify key areas where they can improve employee retention.

## Scenario

Imagine you are working as a data analyst for a mid-sized company that is experiencing high employee turnover, especially among high-performing employees. The company has been facing increased costs related to hiring and training new employees, and management is concerned about the negative impact on productivity and morale. The human resources (HR) team has collected historical employee data and now looks to you for actionable insights. They want to understand why employees are leaving and how to retain talent effectively.

Your task is to analyze the dataset and provide insights that will help HR prioritize retention strategies. These strategies could include interventions like revising compensation policies, improving job satisfaction, or focusing on work-life balance initiatives. The success of your analysis could lead to significant cost savings for the company and an increase in employee engagement and performance.

## Understanding the source of data

The dataset used for this project provides information about employee demographics, performance metrics, and various satisfaction ratings. The dataset is particularly useful for exploring how factors such as job satisfaction, work-life balance, and training opportunities influence employee performance and attrition.

This dataset is well-suited for conducting in-depth analysis of employee performance and retention, enabling us to build predictive models that identify the key drivers of employee attrition. Additionally, we can assess the impact of various organizational factors, such as training and work-life balance, on both performance and retention outcomes.

```{r}
#| label: variable descriptions

## datatable function from DT package create an HTML widget display of the dataset
## install DT package if the package is not yet available in your R environment
read_excel("dataset/dataset-variable-description.xlsx") |> 
  DT::datatable()
```

## Data wrangling

### Data importation

::: callout-tip
## Task 4.1. Merging dataset

-   Import the two dataset `Employee.csv` and `PerformanceRating.csv`. Save the `Employee.csv` as `employee_dta` and `PerformanceRating.csv` as `perf_rating_dta`.

-   Merge the two dataset using the `left_join` function from `dplyr`. Use the `EmployeeID` variable as the varible to join by. You may read more information about the `left_join` function [here](https://dplyr.tidyverse.org/reference/mutate-joins.html).

-   Save the merged dataset as `hr_perf_dta`.
:::

```{r}

## import the two dataset
employee_dta <- read_csv("dataset/Employee.csv")
perf_rating_dta <- read_csv("dataset/PerformanceRating.csv")

## merge employee_dta and perf_rating_dta
hr_perf_dta <- 
  employee_dta |> 
  left_join(perf_rating_dta, by = "EmployeeID") |> 
  mutate(bi_attrition = if_else(Attrition == "No", 0, 1))


## Use the datatable from DT package to display the merged dataset
DT::datatable(hr_perf_dta)

```

### Data management

::: callout-tip
## Task 4.2. Standardizing variable names

-   Using the `clean_names` function from `janitor` package, standardize the variable names by using the recommended naming of variables.

-   Save the renamed variables as `hr_perf_dta` to update the dataset.
:::

```{r}

## clean names using the janitor packages and save as hr_perf_dta
hr_perf_dta <- 
  hr_perf_dta |> 
  clean_names()
```

::: callout-tip
## Recode data entries

-   Create a new variable `cat_education` wherein `education` is `1` = `No formal education`; `2` = `High school`; `3` = `Bachelor`; `4` = `Masters`; `5` = `Doctorate`. Use the `case_when` function to accomplish this task.

-   Similarly, create new variables `cat_envi_sat`, `cat_job_sat`, and `cat_relation_sat` for `environment_satisfaction`, `job_satisfaction`, and `relationship_satisfaction`, respectively. Re-code the values accordingly as `1` = `Very dissatisfied`; `2` = `Dissatisfied`; `3` = `Neutral`; `4` = `Satisfied`; and `5` = `Very satisfied`.

-   Create new variables `cat_work_life_balance`, `cat_self_rating`, `cat_manager_rating` for `work_life_balance`, `self_rating`, and `manager_rating`, respectively. Re-code accordingly as `1` = `Unacceptable`; `2` = `Needs improvement`; `3` = `Meets expectation`; `4` = `Exceeds expectation`; and `5` = `Above and beyond`.

-   Create a new variable `bi_attrition` by transforming `attrition` variable as a numeric variabe. Re-code accordingly as `No` = `0`, and `Yes` = `1`.

-   Save all the changes in the `hr_perf_dta`.
:::

```{r}
## create cat_education


## create cat_envi_sat,  cat_job_sat, and cat_relation_sat


## create cat_work_life_balance, cat_self_rating, and cat_manager_rating


## print the updated hr_perf_dta using datatable function
```

## Exploratory data analysis

### Descriptive statistics of employee attrition

::: callout-tip
## Task 5.1.1.Breakdown of attrition by key variables

-   Select the variables `attrition`, `job_role`, `department`, `age`, `salary`, `job_satisfaction`, and `work_life_balance.` Save as `attrition_key_var_dta`.

-   Compute and plot the attrition rate across `job_role`, `department`, and `age`, `salary`, `job_satisfaction`, and `work_life_balance`.

-   Attrition rate across `job_role` has been done for you! You have the freedom to customize your plot accordingly. Show your creativity!
:::

```{r}
## selecting attrition key variables and save as `attrition_key_var_dta`


## display the attrition_key_var_dta using datatable function.
```

```{r}
#| fig-height: 7

## calculating and plotting attrition rate across job roles
hr_perf_dta |> # replace this data with the attrition_key_var_dta
  group_by(job_role) |> 
  count(attrition) |> 
  mutate(pct_attrition = n / sum(n)) |> 
  ungroup() |> 
  mutate(job_role = reorder_within(job_role, pct_attrition, attrition)) |> 
  ggplot(aes(pct_attrition, job_role, fill = attrition)) +
  geom_col(position = "dodge", width = 0.8) +
  scale_y_reordered() +
  facet_wrap(~ attrition, scales = "free_y", ncol = 1) +
  labs(x = "Attrition rate",
       y = "Job role")
  
```

### Identifying attrition key drivers using correlation analysis

::: callout-tip
## Task 5.2.1. Conduct a correlation analysis to identify key drivers

-   Conduct a correlation analysis of key variables: `bi_attrition`, `salary`, `years_at_company`, `job_satisfaction`, `manager_rating`, and `work_life_balance`. Use the `cor()` function to run the correlation analysis. Remove missing values using the `na.omit()` before running the correlation analysis. Save the output in `hr_corr`.

-   Use a correlation matrix or heatmap to visualize the relationship between these variables and attrition. You can use the `GGally` package and use the `ggcorr` function to visualize the correlation heatmap. You may explore this site for more information: [ggcorr](https://briatte.github.io/ggcorr/).

-   Discuss which factors seem most correlated with attrition and what that suggests aobut why employees are leaving.
:::

```{r}
## conduct correlation of key variables. 
hr_corr <- 
  hr_perf_dta |> 
  select(bi_attrition, salary, years_at_company, job_satisfaction, manager_rating, work_life_balance) |> 
  na.omit() |> 
  cor()

## print hr_corr 
hr_corr

```

```{r}
## install GGally package and use ggcorr function to visualize the correlation


```


### Predictive modeling for attrition

::: callout-tip

## Task 5.3.1. Predictive modeling for attrition

-   Create a logistic regression model to predict employee attrition using the following variables: `salary`, `years_at_company`, `job_satisfaction`, `manager_rating`, and `work_life_balance`. Save the model as `hr_attrition_glm_model`. Print the summary of the model using the `summary` function.

- Install the `sjPlot` package and use the `tab_model` function to display the summary of the model. You may read the documentation [here](https://strengejacke.github.io/sjPlot/articles/tab_model_estimates.html) on how to customize your model summary.

- Also, use the `plot_model` function to visualize the model coefficients. You may read the documentation [here](https://strengejacke.github.io/sjPlot/articles/plot_model_estimates.html) on how to customize your model visualization.

- Discuss the results of the logistic regression model and what they suggest about the factors that contribute to employee attrition.
:::




```{r}
## run a logistic regression model to predict employee attrition
## save the model as hr_attrition_glm_model
hr_attrition_glm_model <- 
  glm(bi_attrition ~ salary + years_at_company + job_satisfaction + manager_rating + work_life_balance, 
      data = hr_perf_dta, 
      family = "binomial")

## print the summary of the model using the summary function
summary(hr_attrition_glm_model)

```

```{r}
## install sjPlot package and use tab_model function to display the summary of the model
library(sjPlot)

tab_model(hr_attrition_glm_model)


```

```{r}
## use plot_model function to visualize the model coefficients
plot_model(hr_attrition_glm_model)

```



### Analysis of compensation and turnover

::: callout-tip

## Task 5.4.1. Analyzing compensation and turnover

- Compare the average monthly income of employees who left the company (`bi_attrition = 1`) and those who stayed (`bi_attrition = 0`). Use the `t.test` function to conduct a t-test and determine if there is a significant difference in average monthly income between the two groups. Save the results in a variable called `attrition_ttest_results`.

- Install the `report` package and use the `report` function to generate a report of the t-test results.

- Install the `ggstatsplot` package and use the `ggbetweenstats` function to visualize the distribution of monthly income for employees who left and those who stayed. Make sure to map the `bi_attrition` variable to the `x` argument and the `salary` variable to the `y` argument.

- Visualize the `salary` variable for employees who left and those who stayed using `geom_histogram` with `geom_freqpoly`. Make sure to facet the plot by the `bi_attrition` variable and apply `alpha` on the histogram plot.

- Provide recommendations on whether revising compensation policies could be an effective retention strategy.

:::

```{r}
## compare the average monthly income of employees who left and those who stayed

attrition_ttest_results <- 
  t.test(salary ~ bi_attrition, data = hr_perf_dta)

```

```{r}
## install the report package and use the report function to generate a report of the t-test results
library(report)

report(attrition_ttest_results)

```

```{r}
# install ggstatsplot package and use ggbetweenstats function to visualize the distribution of monthly income for employees who left and those who stayed
library(ggstatsplot)

ggbetweenstats(data = hr_perf_dta, x = bi_attrition, y = salary)

```


```{r}
# create histogram and frequency polygon of salary for employees who left and those who stayed

hr_perf_dta |> 
  ggplot(aes(x = salary)) +
  geom_histogram(alpha = 0.3) +
  geom_freqpoly() +
  facet_wrap(~ bi_attrition) +
  labs(x = "Attrition",
       y = "Monthly income")
```


### Employee satisfaction and performance analysis

::: callout-tip

## Task 5.5.1. Analyzing employee satisfaction and performance

- Analyze the average performance ratings (both `ManagerRating` and `SelfRating`) of employees who left vs. those who stayed. Use the `group_by` and `count` functions to calculate the average performance ratings for each group.

- Visualize the distribution of `SelfRating` for employees who left and those who stayed using a bar plot. Use the `ggplot` function to create the plot and map the `SelfRating` variable to the `x` argument and the `bi_attrition` variable to the `fill` argument.

- Similarly, visualize the distribution of `ManagerRating` for employees who left and those who stayed using a bar plot. Make sure to map the `ManagerRating` variable to the `x` argument and the `bi_attrition` variable to the `fill` argument.

- Create a boxplot of `salary` by `job_satisfaction` and `bi_attrition` to analyze the relationship between salary, job satisfaction, and attrition. Use the `geom_boxplot` function to create the plot and map the `salary` variable to the `x` argument, the `job_satisfaction` variable to the `y` argument, and the `bi_attrition` variable to the `fill` argument. You need to transform the `job_satisfaction` and `bi_attrition` variables into factors before creating the plot or within the `ggplot` function.

- Discuss the results of the analysis and provide recommendations for HR interventions based on the findings.

:::

```{r}
# Analyze the average performance ratings (both ManagerRating and SelfRating) of employees who left vs. those who stayed.

hr_perf_dta |> 
  na.omit() %>% 
  group_by(bi_attrition) |> 
  summarise(avg_manager_rating = mean(manager_rating, na.rm = TRUE),
            avg_self_rating = mean(self_rating, na.rm = TRUE))

```

```{r}
# Visualize the distribution of SelfRating for employees who left and those who stayed using a bar plot.

hr_perf_dta |> 
  na.omit() %>% 
  ggplot(aes(x = self_rating, fill = factor(bi_attrition))) +
  geom_bar(position = "dodge") +
  labs(x = "Self Rating",
       y = "Count",
       fill = "Attrition")

```

```{r}
# Visualize the distribution of ManagerRating for employees who left and those who stayed using a bar plot.
hr_perf_dta |> 
  na.omit() %>%
  ggplot(aes(x = manager_rating, fill = factor(bi_attrition))) +
  geom_bar(position = "dodge") +
  labs(x = "Manager Rating",
       y = "Count",
       fill = "Attrition")

```


```{r}
# create a boxplot of salary by job_satisfaction and bi_attrition to analyze the relationship between salary, job satisfaction, and attrition.

hr_perf_dta |> 
  na.omit() %>%
  ggplot(aes(y = factor(job_satisfaction), x = salary, fill = factor(bi_attrition))) +
  geom_boxplot() +
  labs(x = "Job Satisfaction",
       y = "Salary",
       fill = "Attrition")

```



### Work-life balance and retention strategies

::: callout-tip
## Task 5.6.1. Analyzing work-life balance and retention strategies

At this point, you are already well aware of the dataset and the possible factors that contribute to employee attrition. Using your R skills, accomplish the following tasks:

- Analyze the distribution of WorkLifeBalance ratings for employees who left versus those who stayed.

- Use visualizations to show the differences.

- Assess whether employees with poor work-life balance are more likely to leave.

You have the freedom how you will accomplish this task. Be creative and provide insights that will help HR develop effective retention strategies.

:::


### Recommendations for HR interventions

::: callout-tip

## Task 5.7.1. Recommendations for HR interventions

Based on the analysis conducted, provide recommendations for HR interventions that could help reduce employee attrition and improve overall employee satisfaction and performance. You may use the following question as guide for your recommendations and discussions.

- What are the key factors contributing to employee attrition in the company?

- Which factors are most strongly correlated with attrition?

- What strategies could be implemented to improve employee retention and satisfaction?

- How can HR leverage the insights from the analysis to develop effective retention strategies?

- What are the potential benefits of implementing these strategies for the company?

:::
  
